En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Supported by organization x.}

### 2019-04-26 mpc19/paper.tex
Types and terms of pi0

\[
A ::= \Z \, | \,A \oplus A \, | \,\I \,| \,A \otimes A \,| \,X \,|
\,\mu X.A 
\]

\begin{figure}
\[
\arraycolsep=8pt\def\arraystretch{2.8}
\begin{array}{ccc}
%\hline 
\infer{\Z : \Ty\,n}{} 
& \infer{\I : \Ty\,n}{}
& \infer{\Var\,i : \Ty\,n}{i \in [0..n)}  \\
\infer{A \oplus B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
& \infer{A \otimes B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
& \infer{\mu A : \Ty\,n}{A : \Ty\,(n+1)} 
%\\ \hline
\end{array}
\]
\caption{Types of \Pio, as formalized in Agda}
\label{fig:types}
\end{figure}

\subsection{Programs}

\begin{figure}
\[
\arraycolsep=8pt\def\arraystretch{1.8}
\begin{array}{cc}%{|cc|}
%\hline 
\id : A \lr A
& \infer{g \bullet f : A \lr C}{g : B \lr C & f : A \lr B} \\ [.25cm]
\infer{f \oplus g : A \oplus B \lr C \oplus D}{f : A \lr C & g : B \lr D} 
& \infer{f \otimes g : A \otimes B \lr C \otimes D}{f : A \lr C & g : B \lr D} \\
\lambda_{\oplus} : \Z \oplus A \lr A
& \lambda_{\oplus}^{-1} : A \lr \Z \oplus A\\
\lambda_{\otimes} : \I \otimes A \lr A
& \lambda_{\otimes}^{-1} : A \lr \I \otimes A\\
\alpha_{\oplus} : (A \oplus B) \oplus C \lr A \oplus (B \oplus C)
& \alpha_{\oplus}^{-1} : A \oplus (B \oplus C) \lr (A \oplus B) \oplus C \\
\alpha_{\otimes} : (A \otimes B) \otimes C \lr A \otimes (B \otimes C)
& \alpha_{\otimes}^{-1} : A \otimes (B \otimes C) \lr (A \otimes B) \otimes C \\
\sigma_{\oplus} : A \oplus B \lr B \oplus A
& \sigma_{\otimes} : A \otimes B \lr B \otimes A \\
\kappa : \Z \otimes A \lr \Z 
& \delta : (A \oplus B) \otimes C \lr (A \otimes C) \oplus (B \otimes C) \\
\kappa^{-1} : \Z \lr \Z \otimes A
& \delta^{-1} : (A \otimes C) \oplus (B \otimes C)  \lr (A \oplus B) \otimes C \\
\fold : A[\mu X.A/X] \lr \mu X.A
& \unfold : \mu X.A \lr A[\mu X.A/X] \\[.25cm]
\multicolumn{2}{c}{\infer{\trace \,f: A \lr B}{f : A \oplus C \lr B \oplus C}} \\
%\hline
\end{array}
\]
\caption{Programs of \Pio}
\label{fig:programs}
\end{figure}

@@

\subsection{Program Equivalences}

\begin{figure}
\[
\arraycolsep=8pt\def\arraystretch{1.8}
\begin{array}{cc} %{|cc|} \hline 
\multicolumn{2}{c}{\mathsf{naturality_L} : f \bullet \trace\,g \LR \trace\,((f \oplus
  \id) \bullet g)} \\ 
\multicolumn{2}{c}{\mathsf{naturality_R} : \trace\,g \bullet f \LR \trace\,(g \bullet (f 
  \oplus \id))} \\  
\multicolumn{2}{c}{\mathsf{dinaturality} : \trace \,((\id \oplus f) \bullet g) \LR \trace
  \,(g \bullet (\id \oplus f))} \\
\multicolumn{2}{c}{\mathsf{vanishing_\oplus} : \trace \,f \LR \trace \,(\trace
  \,(\alpha_\oplus^{-1} \bullet f \bullet \alpha_\oplus))} \\ 
\mathsf{vanishing_\Z} : f \LR \rho_\oplus^{-1} \bullet \trace \,f
  \bullet \rho_\oplus 
&
\mathsf{yanking} : \trace \sigma_\oplus \LR \id \\
\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet (\dagger \,f)
  \bullet g \bullet (\dagger \, g) \LR g \bullet (\dagger \,g)
  \bullet f \bullet (\dagger \, f)} \\
\mathsf{foldIso} : \fold \bullet \unfold \LR \id 
&
\mathsf{unfoldIso} : \unfold \bullet \fold \LR \id
\\[.25cm]
\multicolumn{2}{c}{\infer{\mathsf{tracePIso}\,p : \trace\,f \bullet \trace\,(\dagger\,f)
  \bullet \trace\,f \LR \trace \,f}{p : f \bullet (\dagger\,f) \bullet
  f \LR f}}
%\\ \hline
\end{array}
\]
\caption{Selection of program equivalences of \Pio}
\label{fig:programequivs}
\end{figure}

@@

We now show that $\traceD$ is a dagger trace operator,
i.e. if $f$ is a partial isomorphism, then
$\traceD\,f$ is also a partial isomorphism with partial inverse $\traceD\,(\daggerD\,f)$.

@@

\section{Interpretation}\label{sec:interpretation}

>>>>

The types of \Pio\ are given by the grammar:
\[
A ::= \Z \, | \,A \oplus A \, | \,\I \,| \,A \otimes A \,| \,X \,|
\,\mu X.A 
\]
where $X$ ranges over a set of variables. In Agda, we use de Bruijn
indexes to deal with type variables, so the grammar above is formally
realized by the rules in Figure \ref{fig:types}. The type $\Ty\,n$
represents \Pio\ types containing at most $n$ free
variables. Variables themselves are encoded as elements of
$\mathsf{Fin}\,n$, the type of natural numbers strictly smaller then
$n$. Notice that the $\mu$ binds a variable, which, for $A : \Ty\,(n +
1)$, we consider to be $n + 1$.

\begin{figure}
\[
\arraycolsep=8pt\def\arraystretch{2.8}
\begin{array}{ccc}
%\hline 
\infer{\Z : \Ty\,n}{} 
& \infer{\I : \Ty\,n}{}
& \infer{\Var\,i : \Ty\,n}{i : \mathsf{Fin}\,n}  \\
\infer{A \oplus B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
& \infer{A \otimes B : \Ty\,n}{A : \Ty\,n & B : \Ty\,n}
& \infer{\mu A : \Ty\,n}{A : \Ty\,(n+1)} 
%\\ \hline
\end{array}
\]
\caption{Types of \Pio, as formalized in Agda}
\label{fig:types}
\end{figure}

\subsection{Terms}

The terms of \Pio\ are inductively generated by the rules in Figure
\ref{fig:programs}. They include the identity programs $\id$ and
sequential composition of programs $\bullet$. $(\Z,\oplus)$ is a
symmetric monoidal structure, with terms $\lambda_\oplus$,
$\alpha_\oplus$ and $\sigma_\otimes$ as structural
morphisms. Similarly for $(\I,\otimes)$. Moreover $\otimes$
distributes over $\Z$ and $\oplus$, as evidenced by  $\kappa$
and $\delta$. Elements of $\mu X.A$ are built using the term
constructor $\fold$ and destructed with $\unfold$. Finally, we find the
$\trace$ combinator.

Every \Pio\ program is reversible. The inverse of program is given by
the function $\dagger : (A \lr B) \to (B \lr A)$ recursively defined below:
\begin{figure}
\[
\arraycolsep=7pt\def\arraystretch{1.8}
\begin{array}{cc}%{|cc|}
%\hline 
\id : A \lr A
& \infer{g \bullet f : A \lr C}{g : B \lr C & f : A \lr B} \\ [.25cm]
\infer{f \oplus g : A \oplus B \lr C \oplus D}{f : A \lr C & g : B \lr D} 
& \infer{f \otimes g : A \otimes B \lr C \otimes D}{f : A \lr C & g : B \lr D} \\
\lambda_{\oplus} : \Z \oplus A \lr A
& \lambda_{\oplus}^{-1} : A \lr \Z \oplus A\\
\lambda_{\otimes} : \I \otimes A \lr A
& \lambda_{\otimes}^{-1} : A \lr \I \otimes A\\
\alpha_{\oplus} : (A \oplus B) \oplus C \lr A \oplus (B \oplus C)
& \alpha_{\oplus}^{-1} : A \oplus (B \oplus C) \lr (A \oplus B) \oplus C \\
\alpha_{\otimes} : (A \otimes B) \otimes C \lr A \otimes (B \otimes C)
& \alpha_{\otimes}^{-1} : A \otimes (B \otimes C) \lr (A \otimes B) \otimes C \\
\sigma_{\oplus} : A \oplus B \lr B \oplus A
& \sigma_{\otimes} : A \otimes B \lr B \otimes A \\
\kappa : \Z \otimes A \lr \Z 
& \delta : (A \oplus B) \otimes C \lr (A \otimes C) \oplus (B \otimes C) \\
\kappa^{-1} : \Z \lr \Z \otimes A
& \delta^{-1} : (A \otimes C) \oplus (B \otimes C)  \lr (A \oplus B) \otimes C \\
\fold : A[\mu X.A/X] \lr \mu X.A
& \unfold : \mu X.A \lr A[\mu X.A/X] \\[.25cm]
\multicolumn{2}{c}{\infer{\trace \,f: A \lr B}{f : A \oplus C \lr B \oplus C}} \\
%\hline
\end{array}
\]
\caption{Terms of \Pio}
\label{fig:programs}
\end{figure}

\[
\begin{array}{llcll}
%\multicolumn{2}{l}{\dagger : \{A B : \Ty \,n\} \to A \lr B \to B \lr A} \\
\dagger \,\id &= \id & \quad &
\dagger\, (g \bullet f) & = \dagger\, f \bullet \dagger\, g \\
\dagger\, (f \oplus g) & = \dagger\, f \oplus \dagger\, g & \quad &
\dagger\, (f \otimes g) & = \dagger\, f \otimes \dagger\, g \\
\dagger\, \lambda_\oplus ^{-1} & = \lambda_\oplus & \quad &
\dagger\, \lambda_\oplus & = \lambda_\oplus^{-1} \\
\dagger\, \sigma_\oplus & = \sigma_\oplus & \quad &
\dagger\, \alpha_\oplus & = \alpha_\oplus^{-1} \\
\dagger\, \alpha_\oplus^{-1} & = \alpha_\oplus & \quad &
\dagger\, \lambda_\otimes & = \lambda_\otimes^{-1} \\
\dagger\, \lambda_\otimes ^{-1} & = \lambda_\otimes & \quad &
\dagger\, \sigma_\otimes & = \sigma_\otimes \\
\dagger\, \alpha_\otimes & = \alpha_\otimes^{-1} & \quad &
\dagger\, \alpha_\otimes^{-1} & = \alpha_\otimes \\
\dagger\, \kappa & = \kappa^{-1} & \quad &
\dagger\, \kappa^{-1} & = \kappa \\
\dagger\, \delta & = \delta^{-1} & \quad &
\dagger\, \delta^{-1} & = \delta \\
\dagger\, \fold & = \unfold & \quad &
\dagger\, \unfold & = \fold \\
\dagger\, (\trace \,f) & = \trace \,(\dagger\, f)
\end{array}
\]

@@

\subsection{Term Equivalences}

\begin{figure}
\[
\arraycolsep=8pt\def\arraystretch{1.8}
\begin{array}{cc} %{|cc|} \hline 
\multicolumn{2}{c}{\mathsf{naturality_L} : f \bullet \trace\,g \LR \trace\,((f \oplus
  \id) \bullet g)} \\ 
\multicolumn{2}{c}{\mathsf{naturality_R} : \trace\,g \bullet f \LR \trace\,(g \bullet (f 
  \oplus \id))} \\  
\multicolumn{2}{c}{\mathsf{dinaturality} : \trace \,((\id \oplus f) \bullet g) \LR \trace
  \,(g \bullet (\id \oplus f))} \\
\multicolumn{2}{c}{\mathsf{vanishing_\oplus} : \trace \,f \LR \trace \,(\trace
  \,(\alpha_\oplus^{-1} \bullet f \bullet \alpha_\oplus))} \\ 
\mathsf{vanishing_\Z} : f \LR \rho_\oplus^{-1} \bullet \trace \,f
  \bullet \rho_\oplus 
&
\mathsf{yanking} : \trace \sigma_\oplus \LR \id \\
\multicolumn{2}{c}{\mathsf{uniquePIso} : f \bullet \dagger \,f
  \bullet g \bullet \dagger \, g \LR g \bullet \dagger \,g
  \bullet f \bullet \dagger \, f} \\
\mathsf{foldIso} : \fold \bullet \unfold \LR \id 
&
\mathsf{unfoldIso} : \unfold \bullet \fold \LR \id
\\[.25cm]
\multicolumn{2}{c}{\infer{\mathsf{tracePIso}\,p : \trace\,f \bullet \trace\,(\dagger\,f)
  \bullet \trace\,f \LR \trace \,f}{p : f \bullet \dagger\,f \bullet
  f \LR f}}
%\\ \hline
\end{array}
\]
\caption{Selection of term equivalences of \Pio}
\label{fig:programequivs}
\end{figure}

@@

We now move to show that $\traceD$ is a dagger trace operator,
i.e. if $f$ is a partial isomorphism, then
$\traceD\,f$ is also a partial isomorphism with partial inverse $\traceD\,(\daggerD\,f)$.

@@

\section{Soundness}\label{sec:interpretation}
